Noun Serialization

### 2025-11-05 mss.tex
Realign +fat

\newcommand{\authorname}{N. E. Davis, Brian Klatt, \& Sam Parker}
\newcommand{\authorpatp}{\patp{lagrev-nocfep}, \patp{zod}, \& \patp{zod}}
\newcommand{\affiliation}{Zorp Corp, Zorp Corp, \& —}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} II:2 (2025):  1–46. \\
              % ~ \\
              % Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Serializing Nouns}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{Serializing Nouns}
\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{+jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{+bulk}. \\ We gratefully acknowledge the contributions of the original architects of \texttt{+jam}/\texttt{+cue} as well.} \\ \affiliation}
% TODO Brian Klatt credit
\date{}

@@

\begin{abstract}
Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.  This article describes and compares the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces variant encodings for educational purposes.
\end{abstract}

@@

A noun serialization strategy is rather like a Gödel numbering in that it systematically encodes a mathematical object (a noun) as a number (an atom).  Unlike Gödel numbering, which classically serially encodes the symbols of mathematical statements, noun serialization encodes a binary tree structure.\footnote{This also echoes the way that S-expressions are encoded in Lisp.}  Because a noun may be any atom---and atoms cannot have leading zeroes---both structure and value need to be unambiguously encoded and cannot be simply delimited (as by a \texttt{0} bit or similar).  There are two basic strategies to encode a noun as an atom:

\begin{enumerate}
  \item  Run-length serialization, with or without references.
  \item  Directed graph serialization, depending on a reentrant graph encoding.
\end{enumerate}

\noindent
Both of these embody a tradeoff between simplicity, size, and speed.  Below, we describe both of these strategies and some new variants which offer possible advantages.  Reversible noun encoding is essential for Nock communication, both between Nock execution layers such as Vere and NockVM, and between the runtime and the host operating system.

@@

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).}
\label{tab:atom-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b110} \\
  \texttt{0b10} & 2 & \texttt{0b10100} \\
  \texttt{0b11} & 2 & \texttt{0b11100} \\
  \texttt{0b101} & 3 & \texttt{0b1011000} \\
  \texttt{0b111} & 3 & \texttt{0b1111000} \\
  \texttt{0x70} & 7 & \texttt{0b111000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
\end{tabular}
\end{table}

\noindent
We can make one further optimization:  since the most significant bit (\textsc{msb}) of the atom is always \texttt{1} except in the case of zero (which is \texttt{0b10} with the leading zero trimmed), we can omit it from the encoded form.  This saves one bit for every non-zero atom.  Let's call this scheme \texttt{+fat} encoding.  (See Table~\ref{tab:fat-encodings} for examples.)

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings (\texttt{+fat} encoding).}
\label{tab:fat-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b10} \\
  \texttt{0b10} & 2 & \texttt{0b100} \\
  \texttt{0b11} & 2 & \texttt{0b1100} \\
  \texttt{0b101} & 3 & \texttt{0b11000} \\
  \texttt{0b111} & 3 & \texttt{0b111000} \\
  \texttt{0x70} & 7 & \texttt{0b11000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b1111111100000000} \\
\end{tabular}
\end{table}

@@

Consider the noun \texttt{[a [b c]]}, represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:

@@

\begin{lstlisting}[style=listingcode]
++  jam
  ~/  %jam
  |=  a=*
  ^-  @
  =+  b=0
  =+  m=`(map * @)`~
  =<  q
  |-  ^-  [p=@ q=@ r=(map * @)]
  =+  c=(~(get by m) a)
  ?~  c
    =>  .(m (~(put by m) a b))
    ?:  ?=(@ a)
      =+  d=(mat a)
      [(add 1 p.d) (lsh 0 q.d) m]
    =>  .(b (add 2 b))
    =+  d=$(a -.a)
    =+  e=$(a +.a, b (add b p.d), m r.d)
    :+  (add 2 (add p.d p.e))
      (mix 1 (lsh [0 2] (cat 0 q.d q.e)))
    r.e
  ?:  ?&(?=(@ a) (lte (met 0 a) (met 0 u.c)))
    =+  d=(mat a)
    [(add 1 p.d) (lsh 0 q.d) m]
  =+  d=(mat u.c)
  [(add 2 p.d) (mix 3 (lsh [0 2] q.d)) m]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
TODO
\end{lstlisting}

@@

\section{Directed Graph Encoding}

A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{+bulk} encoding.

@@

\section{Benchmarks}

@@

\section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}

The following is a simple Python implementation of \texttt{+jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.

>>>>

\newcommand{\authorname}{Brian Klatt \patp{ticlys-monlun}, N. E. Davis \patp{lagrev-nocfep}}
\newcommand{\affiliation}{Zorp Corp}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} II:2 (2025):  1–46. \\
              % ~ \\
              % Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Noun Serialization}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{Noun Serialization}
\author{Brian Klatt \patp{ticlys-monlun}, \\ N. E. Davis \patp{lagrev-nocfep}* \\ \affiliation}
\date{}

@@

\begin{abstract}
Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.  This article describes the principal convention for representing a noun in compressed form as a byte array, as well as introduces variant encodings for educational purposes.
\end{abstract}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{0}
\footnotetext[1]{\sloppy We gratefully acknowledge the original architect of \texttt{+jam} and \texttt{+cue}, \mbox{Curtis} Yarvin \patp{sorreg-namtyv}.  The original function, in \href{https://github.com/cgyarvin/urbit/blame/55f591594080d276fc446ceafa4afff092561613/watt/jam.watt}{\texttt{jam.watt}}, dates to \textasciitilde 2011.6.11.}
\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

@@

A noun serialization strategy is rather like a Gödel numbering in that it systematically encodes a mathematical object (a noun) as a number (an atom).  Unlike Gödel numbering, which classically serially encodes the symbols of mathematical statements, noun serialization encodes a binary tree structure.\footnote{This also echoes the way that S-expressions are encoded in Lisp.}  Because a noun may be any atom---and atoms cannot have leading zeroes---both structure and value need to be unambiguously encoded and cannot be simply delimited (as by a \texttt{0} bit or similar).  Run-length serialization, with or without references, forms the basis of practical strategies to encode a noun as an atom.  This embodies a tradeoff between simplicity, size, and speed.  We describe this strategy and some pedagogical variants which afford insight to the process.  Likewise, deserialization is the inverse operation of converting a serialized atom back into the original noun, and must be unambiguously reversible.  Noun serialization is essential for Nock communication, both between Nock execution layers such as Vere and NockVM, and between the runtime and the host operating system.

@@

\subsection{Atom Length Encoding}

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).}
\label{tab:atom-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b110} \\
  \texttt{0b10} & 2 & \texttt{0b10100} \\
  \texttt{0b11} & 2 & \texttt{0b11100} \\
  \texttt{0b101} & 3 & \texttt{0b1011000} \\
  \texttt{0b111} & 3 & \texttt{0b1111000} \\
  \texttt{0x70} & 7 & \texttt{0b111000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
  \texttt{0x1234} & 15 & \texttt{0b100001100100001} \\
  & & \texttt{↳1000000000000000} \\
\end{tabular}
\end{table}

\noindent
We can make one further optimization:  since the most significant bit (\textsc{msb}) of the atom is always \texttt{1} except in the case of zero (which is \texttt{0b1} with the leading zero trimmed), we can omit it from the encoded form.  This saves one bit for every non-zero atom.  Let's call this scheme \texttt{+fat} encoding.  (See Table~\ref{tab:fat-encodings} for examples.)

@@

\begin{table}[ht]
\centering
\caption{Example atom encodings (\texttt{+fat} encoding).  Elided digits are shown in light gray for clarity.}
\label{tab:fat-encodings}
\begin{tabular}{ l l r }
  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
  \hline
  \texttt{0b0} & 1 & \texttt{0b1} \\
  \texttt{0b1} & 1 & \texttt{0b\textcolor{lightgray}{1}10} \\
  \texttt{0b10} & 2 & \texttt{0b\textcolor{lightgray}{10}100} \\
  \texttt{0b11} & 2 & \texttt{0b\textcolor{lightgray}{1}1100} \\
  \texttt{0b101} & 3 & \texttt{0b\textcolor{lightgray}{10}11000} \\
  \texttt{0b111} & 3 & \texttt{0b\textcolor{lightgray}{1}111000} \\
  \texttt{0x70} & 7 & \texttt{0b\textcolor{lightgray}{1}11000010000000} \\
  \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
  \texttt{0x1234} & 15 & \texttt{0b\textcolor{lightgray}{10000}1100100001} \\
  & & \texttt{↳1000000000000000} \\
\end{tabular}
\end{table}

@@

Consider the noun \texttt{[a [b c]]} (= \texttt{[a b c]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:

@@

\begin{lstlisting}[style=listingcode]
++  jam
  ~/  %jam
  |=  a=*
  ^-  @
  =+  b=0
  =+  m=`(map * @)`~
  =<  q
  |-  ^-  [p=@ q=@ r=(map * @)]
  =+  c=(~(get by m) a)
  ?~  c
    =>  .(m (~(put by m) a b))
    ?:  ?=(@ a)
      =+  d=(mat a)
      [(add 1 p.d) (lsh 0 q.d) m]
    =>  .(b (add 2 b))
    =+  d=$(a -.a)
    =+  e=$(a +.a, b (add b p.d), m r.d)
    :+  (add 2 (add p.d p.e))
      (mix 1 (lsh [0 2] (cat 0 q.d q.e)))
    r.e
  ?:  ?&(?=(@ a) (lte (met 0 a) (met 0 u.c)))
    =+  d=(mat a)
    [(add 1 p.d) (lsh 0 q.d) m]
  =+  d=(mat u.c)
  [(add 2 p.d) (mix 3 (lsh [0 2] q.d)) m]
::  length encode
++  mat
  ~/  %mat
  |=  a=@
  ^-  [p=@ q=@]
  ?:  =(0 a)
    [1 1]
  =+  b=(met 0 a)
  =+  c=(met 0 b)
  :-  (add (add c c) b)
  (cat 0 (bex c) (mix (end [0 (dec c)] b) (lsh [0 (dec c)] a)))
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
++  cue
  ~/  %cue
  |=  a=@
  ^-  *
  =+  b=0
  =+  m=`(map @ *)`~
  =<  q
  |-  ^-  [p=@ q=* r=(map @ *)]
  ?:  =(0 (cut 0 [b 1] a))
    =+  c=(rub +(b) a)
    [+(p.c) q.c (~(put by m) b q.c)]
  =+  c=(add 2 b)
  ?:  =(0 (cut 0 [+(b) 1] a))
    =+  u=$(b c)
    =+  v=$(b (add p.u c), m r.u)
    =+  w=[q.u q.v]
    [(add 2 (add p.u p.v)) w (~(put by r.v) b w)]
  =+  d=(rub c a)
  [(add 2 p.d) (need (~(get by m) q.d)) m]
::  length decode
++  rub
  ~/  %rub
  |=  [a=@ b=@]
  ^-  [p=@ q=@]
  =+  ^=  c
      =+  [c=0 m=(met 0 b)]
      |-  ?<  (gth c m)
      ?.  =(0 (cut 0 [(add a c) 1] b))
        c
      $(c +(c))
  ?:  =(0 c)
    [1 0]
  =+  d=(add a +(c))
  =+  e=(add (bex (dec c)) (cut 0 [d (dec c)] b))
  [(add (add c c) e) (cut 0 [(add d (dec c)) e] b)]
\end{lstlisting}

@@

@@

@@

\tombstone

\selectlanguage{USenglish}
% \printbibliography

\section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}
\subsection*{Paul Driver \patp{fodwyt-ragful}}

The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}
